---
title: "Lab Task 7"
author: "Group 6"
date: Sys.Date()
output:
  html_document:
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---

Group members:

- Shawn Yakir , 315714980,  shawn.yakir@mail.huji.ac.il
- Sarah Levitz, 324673623, sarah.levitz@mail.huji.ac.il
- Carmel Baris, 318455276, carmel.baris@mail.huji.ac.il 

### Clean the environment
```{r clean}
rm(list = ls())
```

### Libraries used

```{r libraries}
library('splines')
library('scales')
```


### Paths
```{r paths}
#path = "C:/Users/user/Documents/Carmel/Projects/HUJI/Archive_HUJI/data-analysis-genomics/data"
path = "/Users/alevi/Downloads/"
# path = "\\Users\\shawn\\Desktop\\"
```


### Helper functions
```{r helpers}
helpers_file = file.path(path, "help_funcs.R")
source(helpers_file)
helpers = list(loadRData = loadRData, binReads = binReads)
```


### Load data
```{r dat}
load(sprintf("%s/reads_100_A2.rda",path))
load(sprintf("%s/reads_100_B2.rda",path))
load(sprintf("%s/GC_100.rda",path))
GC_100=GC_100[1: (length(reads_100_A2)) ]
```

### Binning
```{r binning}
binsize = 25

# Trim the data vectors to the nearest multiple of binsize
trim_length = length(reads_100_A2) - (length(reads_100_A2) %% binsize)
reads_100_A2_trimmed = reads_100_A2[1:trim_length]
reads_100_B2_trimmed = reads_100_B2[1:trim_length]
GC_100_trimmed = GC_100[1:trim_length]

# Create the matrices
chr2t_mat = matrix(reads_100_A2_trimmed, nrow = binsize)
reads2.5K_t2 = colSums(chr2t_mat)

chr2n_mat = matrix(reads_100_B2_trimmed, nrow = binsize)
reads2.5K_n2 = colSums(chr2n_mat)

gc_mat = matrix(GC_100_trimmed, nrow = binsize)
GC2.5K = colMeans(gc_mat)
```


### Outliers
```{r outliers}
# Outliers of GC percent per bin
zeros_GC = GC2.5K == 0
# Identify outliers of reads count per bin
z_scores_n2 = (reads2.5K_n2 - mean(reads2.5K_n2)) / sd(reads2.5K_n2)
z_scores_t2 = (reads2.5K_t2 - mean(reads2.5K_t2)) / sd(reads2.5K_t2)
xtreme_reads_n2 = abs(z_scores_n2) > 3
xtreme_reads_t2 = abs(z_scores_t2) > 3

# Create new variables without outliers
outliers = (zeros_GC | xtreme_reads_t2 | xtreme_reads_n2)
outliers = (outliers | (reads2.5K_t2<0.05) | (reads2.5K_n2<0.05))
reads2.5K_n2_clean = reads2.5K_n2[!outliers]
reads2.5K_t2_clean = reads2.5K_t2[!outliers]
GC2.5K_clean = GC2.5K[!outliers]

```

### Fitting bspline regression
```{r fit}
create_est_f = function(X, Y, deg, knots) {
  lknot = min(X)
  rknot = max(X)
  lm(Y ~ bs(X, degree = deg, knots = knots, Boundary.knots = c(lknot, rknot)))
}
```



```{r normalizing}
# Use the rescale function from the scales package
reads2.5K_n2_normalized = scales::rescale(reads2.5K_n2_clean, to = c(0, 100))
reads2.5K_t2_normalized = scales::rescale(reads2.5K_t2_clean, to = c(0, 100))
```

```{r}
# Ensure sample size is not larger than the length of the cleaned data
sample_size = min(length(reads2.5K_n2_clean), 10000)
samp1 = sample(length(reads2.5K_n2_clean), sample_size)

# Plot the normalized data
par(mfcol = c(2, 1), mar = c(1, 5, 1, 1))
plot(GC2.5K_clean[samp1], reads2.5K_n2_normalized[samp1], cex = 0.4, pch = 20, ylab = "Healthy 2", ylim = c(1.5, 35), xlim = c(0.2, 0.75))
plot(GC2.5K_clean[samp1], reads2.5K_t2_normalized[samp1], cex = 0.4, pch = 20, ylab = "Tumor 2", ylim = c(0, 35), xlim = c(0.2, 0.75))

```


## Learn parameters from chromosome 1

```{r}
l_2n = length(reads2.5K_n2_normalized)
l_2t = length(reads2.5K_t2_normalized)

myknots = c(0.415, 0.45, 0.54)

GC=GC2.5K_clean[1:l_2n]

mod_2n = lm(reads2.5K_n2_normalized~bs(GC, degree = 3),subset = reads2.5K_n2_normalized>1.5 & reads2.5K_n2_normalized<35 )

pred_range = seq(0.2,0.75,0.01)
resp_n = predict(mod_2n,list(GC=pred_range))
preds2n = predict(mod_2n,list(GC=GC2.5K_clean))

par(mfcol = c(1,1),mar = c(3,5,3,3))
plot(GC2.5K_clean[samp1],reads2.5K_n2_normalized[samp1],ylim = c(0,35),xlim = c(0.2,.75), cex = 0.5, pch=20, main = "Healthy 2")
lines(pred_range, resp_n,col = 4,lw=3)
lines(pred_range, resp_n,col = 4,lw=3)

```


```{r}
GC=GC2.5K_clean[1:l_2t]
mod_2t = lm(reads2.5K_t2_normalized~bs(GC, degree = 3),subset = (reads2.5K_t2_normalized>1.5)&(reads2.5K_t2_normalized<40))

pred_range = seq(0.2,0.75,0.01)
resp_t = predict(mod_2t,list(GC=pred_range))
preds2t = predict(mod_2t,list(GC=GC2.5K_clean))

par(mfcol = c(1,1),mar = c(3,3,3,3))
plot(GC2.5K_clean[samp1],reads2.5K_t2_normalized[samp1],ylim = c(0,35),xlim = c(0.2,.75), cex = 0.5, pch=20, main = "Tumor 2")
lines(pred_range, resp_t,col = 2,lw=3)
lines(pred_range, resp_n,col = 4,lw=3)
```


## One sample corrections
```{r func_fix}
one_sample_correct = function(samp, preds) {
  one_correct = (samp)/(preds)
  return(one_correct)
}
```

```{r med_fix}
corrected_2n = one_sample_correct (reads2.5K_n2_normalized,preds2n[1:l_2n])
uncorrected_2n = (reads2.5K_n2_normalized)/median(reads2.5K_n2_normalized) 

corrected_2t = one_sample_correct (reads2.5K_t2_normalized,preds2t[1:l_2t])
uncorrected_2t = (reads2.5K_t2_normalized)/median(reads2.5K_t2_normalized) 

```

### Introduction 

This week we...

### Part 1 Q1: 

We'll start with a breakdown of the variance in the healthy sample after filtering out the outliers and after correcting the median.

```{r}
# Fit B-spline Model
l_2n = length(reads2.5K_n2_clean)
GC = GC2.5K_clean[1:l_2n]
mod_2n = lm(reads2.5K_n2_clean ~ bs(GC, degree = 3))
preds2n = predict(mod_2n, list(GC = GC))
residuals_2n = reads2.5K_n2_clean - preds2n
```

```{r}
### Median Correction
uncorrected_2n = reads2.5K_n2_clean / median(reads2.5K_n2_clean)

### Variance Decomposition
total_variance = var(reads2.5K_n2_clean)
poisson_variance = mean(reads2.5K_n2_clean)
gc_variance = var(preds2n)
unexplained_variance = var(residuals_2n)

### Percentage of Variance
percentage_poisson = (poisson_variance / total_variance) * 100
percentage_gc = (gc_variance / total_variance) * 100
percentage_unexplained = (unexplained_variance / total_variance) * 100
```

```{r}
### Results
cat("Percentage of variance due to Poisson distribution: ", percentage_poisson, "%\n")
cat("Percentage of variance due to GC effect: ", percentage_gc, "%\n")
cat("Percentage of variance unexplained: ", percentage_unexplained, "%\n")
```


```{r}
### One sample corrections
one_sample_correct = function(samp, preds) {
  one_correct = (samp) / (preds)
  return(one_correct)
}

# Apply correction function to normalized data
corrected_2n = one_sample_correct(reads2.5K_n2_normalized, preds2n[1:l_2n])
uncorrected_2n = (reads2.5K_n2_normalized) / median(reads2.5K_n2_normalized)

corrected_2t = one_sample_correct(reads2.5K_t2_normalized, preds2t[1:l_2n])
uncorrected_2t = (reads2.5K_t2_normalized) / median(reads2.5K_t2_normalized)

### Calculate Variance for Corrected Data
corrected_total_variance = var(corrected_2n)
corrected_poisson_variance = mean(corrected_2n)
corrected_gc_variance = var(preds2n)
corrected_unexplained_variance = var(corrected_2n - preds2n)

# Calculate percentages correctly
corrected_percentage_poisson = (corrected_poisson_variance / corrected_total_variance) * 100
corrected_percentage_gc = (corrected_gc_variance / corrected_total_variance) * 100
corrected_percentage_unexplained = (corrected_unexplained_variance / corrected_total_variance) * 100

#!!!!!!!!!!!!!!!!!!!!!1 issue with corrected data
### Results for Corrected Data
cat("Percentage of variance due to Poisson distribution (corrected): ", corrected_percentage_poisson, "%\n")
cat("Percentage of variance due to GC effect (corrected): ", corrected_percentage_gc, "%\n")
cat("Percentage of variance unexplained (corrected): ", corrected_percentage_unexplained, "%\n")

```



Poisson Distribution: 6.44% of the total variance is due to the inherent randomness in read counts.
GC Content: 64.64% of the variance is explained by GC content, indicating a strong influence on read counts.
Unexplained Variance: 35.36% of the variance is not explained by the Poisson distribution or GC content, suggesting other factors are at play.


### Part 2 Q1: 

```{r}
# Fit B-spline Regression Models
GC_healthy = GC2.5K_clean
mod_healthy = lm(reads2.5K_n2_clean ~ bs(GC_healthy, degree = 3))
pred_healthy = predict(mod_healthy, list(GC_healthy = GC_healthy))

# Tumor Sample
reads2.5K_t2_clean = reads2.5K_t2_clean[1:l_2n]  # Ensure tumor data matches in length
mod_tumor = lm(reads2.5K_t2_clean ~ bs(GC_healthy, degree = 3))
pred_tumor = predict(mod_tumor, list(GC_healthy = GC_healthy))

```

```{r}
# Define the range for predictions
pred_range = seq(0.2, 0.75, 0.01)

# Predict values over the range for both models
preds_healthy = predict(mod_healthy, newdata = data.frame(GC_healthy = pred_range))
preds_tumor = predict(mod_tumor, newdata = data.frame(GC_healthy = pred_range))

```

```{r}
# Combined Plot
par(mfcol = c(1, 1), mar = c(5, 5, 4, 2))

# Plot the data points first
plot(GC_healthy, reads2.5K_n2_clean, col = 'blue', pch = 20, cex = 0.5, ylab = "Normalized Reads", xlab = "GC Content", main = "Healthy vs Tumor Sample")
points(GC_healthy, reads2.5K_t2_clean, col = 'red', pch = 20, cex = 0.5)

# Add the regression lines on top
lines(pred_range, preds_healthy, col = 'darkblue', lwd = 2)
lines(pred_range, preds_tumor, col = 'darkred', lwd = 2)

# Add legend
legend("topright", legend = c("Healthy Sample", "Tumor Sample"), col = c("blue", "red"), lty = 1, lwd = 2, pch = 20)

```

Shape of the Regression Lines: Both samples exhibit a similar unimodal relationship with GC content, indicating that the GC effect on read counts is similar in shape for both healthy and tumor samples.
Magnitude of Read Counts: The tumor sample generally has higher read counts compared to the healthy sample across the GC content range. This indicates that while the pattern is similar, the extent to which GC content affects read counts differs between the two samples.

The connection is the same in terms of the pattern of the relationship, but the magnitude of the effect differs between the healthy and tumor samples.



