---
title: "formulas_010724"
author: "Carmel Baris"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("splines")
```

So far we have learned that the distribution of $Y_k$, the coverage of bin $K$, depends on the GC content in that bin. We have assumed a general non-parametric regression model of the form $Y \sim f(GC) + \mathcal{E}$. \n


$$
i \text{ is the location within the chromosome being sequenced}\\
i = 1,2,\cdots,I \\
Y_i = \text{Count of Reads in location } I\\
\Omega(Y_i) = \big\{0,\cdots,J\big\} \\
X_i=\text{GC content (%) in location } I\\
\Omega(X_i) = \big\{0,\cdots,1\big\} \\
j \ \text{ is the serial number of a given Read that starts a sequencing fragment}\\
j = 0,2,\cdots,J \\
L_j \ \text{ is the chromosomal location of the } \ j^{\ th} \ \text{ Read}\\
\text{In other words, } L_j \text{ maps each } j^{\ th} \text{ Read to its location } i:\\
\Omega(L_j) = \{ {i} \}\\
Y_i = \sum_{i=1}^{I} \big[𝟙_{L_j=i} \big] \\

\text{For a given location } i \text{ (bin size = 1), we assume that} \  \\
\text{ the regression model is of the non-parametric form: } \\
Y \sim f(X) + \mathcal{E} \ \ \\
\text{where } f(X) \text{ is the regression function that ties the number of reads (Y) to GC-content (X).} \\ 

$$

```{r}
par(mfrow=c(2,1))
plot(reads_2.5K_n1, main = "Marginal Distr. of Read Count in Normative Sample", ylim=c(0,600))
plot(reads_2.5K_t1, main = "Marginal Distr. of Read Count in Tumor Sample", ylim=c(0,600))
plot(GC_2.5K, main = "Marginal Distribution of GC Content (%)")
```

### Marginal distribution of Y_K
$$
B \text{ is a constant that represents the bin size. (e.g, 1 bin : 2,500 bases)} \\
k \in K \text{ is the serial number of a given bin: } K = 1,2,\cdots,\frac{\ I}{B} \\
X_k=\text{GC Content (%) in bin K}; \ 
\Omega(X_K) = \big\{0,\cdots,1\big\} \\
Y_k = \text{Count of Reads in bin k}; \ \Omega(Y_K) = \big\{0,\cdots,J\big\} \\

$$


### Regression  
$$
\textbf{Regression model is of the form: } 
Y \sim f(X) + \mathcal{E} \\
Y_K \sim f(X_K) + \mathcal{E}_K  \\
Y_K \sim f(X_K) + \mathcal{E}_K \\
\hat{Y}_K \sim \hat{f}(GC_K) \\
f(X) = \overline{Y_K} \sim f(GC_K) + \mathcal{E}  \\

\text{To estimate } f(X) \text{ we apply the} \textbf{ Method of Moments (MoM).} \\
\text{First, we define the dependent variable } f(X) \ \text{ to be a conditional expectation:} \\
f(X) = E[Y|X=x] \\
\text{Second, we substitute the conditional expectation with the sample average:} \\
\hat{f}(X) = \hat{E}[Y|X=x] = E[\hat{Y}|X=x] \\
\text{Third, we apply OLS by which the line of mean values converges with the regression line. }
$$

# Piecewise Estimation of f(GC) using BSpline
To estimate the regression line, we use a B-Spline function, 
that provides us with a "smooth" continuous piecewise model:
$$
\hat{f}(X) = lm \bigg( Y_K \sim bspline(X_K, 3^{rd} \text{ degree, knots = } t_1,t_2,t_3, \dots ) \bigg)\\

$$


```{r}
plot(GC_2.5K, reads_2.5K_n1, ylim = c(0,400))
lines(GC_2.5K, lm(reads_2.5K_n1~bs(GC_2.5K),subset = !outli))
```


# Piecewise Estimation of f(GC) using BSpline
## Lesson 7
We like to talk about the *root* of Mean Squared Errors (RMSE), rather than MSE.
$$RMSE( \{ \hat a_k \},1) = \sqrt(\frac{1}{b} \sum_{k =1}^{I} \big(\hat{a_k} - 1\big)^2 \\
EMSE = E_Y \Big[ MSE( \{ \hat a_i \},1) = \sqrt(\frac{1}{b} \sum_{i=1}^{I} \big(\hat{a_i} - 1\big)^2 \\
$$
Same goes when discussing variance. We prefer to use its root, the standard deviation, as it has significance in regards to deviations from the mean expected value.\n


$$
\begin{equation}
\text{Var}[X] \overset{\text{def}}{=} E\Big[\big(X - E[X]\big)^2\Big].
\tag{28.1}
\end{equation}
$$



### fff

Picture the following scenario.
An oncology specialist comes to our lab and wants to know if her patient has cancer.
She gives us a sample of the patient's DNA, from the chromosome in question.
Our task is to categorize this DNA sample as either healthy ("Normative") or cancerous ("Tumor").
It would have been great to have another sample of the very same patient,
 from the exact same cellular tissue that was once healthy and is now cancerous. 
Why? Because then we could compare the count of reads and if we don't get a ratio of 1:1, 
  we assume that this may indicate that the sample is cancerous.
Let's write this a bit more formally:

$$ 
\textbf{What do we expect, if sample } s \textbf{ is healthy?} \\
    \frac{Y^s}{Y^n} =
\bigg(
    \frac{\text{Num. of Reads from Sample }s}
         {\text{Num. of Reads from Normative Sample}} 
            \bigg) 
    \overset{\checkmark}{=} 1
    \\
$$


Unfortunately, that's not an option. 
Instead, we can use  statistical tools to estimate the number of reads in a healthy sample.
The inherent problem of estimation is that there will always be a gap, caused by an unknown factor $\mathcal{E}$.
Our aim is to minimize the size of that epsilon s.t. it doesn't affect our estimation too much.

To do this, we need to better understand what influences the $Y$, i.e. the Read Count in a given sample.
In other words, we want to formulate the distribution function of $Y$.

*First Step*
By looking at the marginal distr. of Read Counts, just for the healthy sample, we see that the Read Count doesn't stay the same.
Rather it changes for different indices. 
Since each index represents a location, we understand that the Read Count depends on the location. 
For example, around the center of the chromosome we see a big drop in the Read Count.
This means that we have very little reads mapped to those locations.
The same holds when we look at the Read Count after binning the data into groups of 2.5K bases (below).

```{r marginal_y}
plot(reads_2.5K_n1, main = "Marginal Distr. of Read Counts (Healthy Sample)", ylim=c(0,600),
     ylab = "Number of reads in bin number K",
     xlab = "Location of bin number K (in Chr1)"
     )
```

So if someone asks us how many reads is considered "healthy" in bin K, my educated guess would be the expected mean value----?.
If the sample is indeed healthy, I will expect:

$$ 
\textbf{What do we expect, if sample s is healthy?} \\
    \frac{Y^s}{Y^n} 
      \propto \frac{Y^s}{{\mathbb{E_k}[Y^n]}} 
      \approx \frac{Y^s}{{\mathbb{\hat{E_k}}[Y^n]}} 
      =\frac{Y^s}{\operatorname{median}(Y^n)} 
    \overset{?}{=} 1
    \\
$$


$$ 
\textbf{What do we expect, if sample s is healthy?} \\
    \frac{Y^s}{Y^n} 
      \approx \frac{Y^s}{{\mathbb{E}[Y^n]}} 
      \approx \frac{Y^s}{{\mathbb{\hat{E}}[Y^n]}} 
      =\frac{Y^s}{\operatorname{median}(Y^n)} 
    \overset{?}{=} 1
    \\
$$



$$
\text{we expect to see changes depending on the location of the bin.} \\
\text{Supposedly } Y_K \sim \text{Location of Bin K} \text{, but we know this is not true!} \\
\text{If we take GC content into account, the actual read count in bin K is } \ Y_K \sim f(\text{GC}{_{K}}) \cdot \mathcal{E}_{K} \\
\text{And then our estimated read count in bin K is  } \ \hat{Y}_K \sim \hat{f}(\text{GC}_{K})  \\
$$


$$
\text{}
$$

### Lab 6
$$
{Y_k}^n \ \text{ Normative sample (Healthy)} \\
{Y_k}^t \sim  \ \text{ Tumor sample (Cancerous)} \\
\lambda_k \ \text{Mean Expected value of number of reads} \\
\mathcal{E}_K \\
\gamma^n \neq \gamma^t \ \text{ Poisson "noise"} \\
\eta^j \ \text{ Joint "noise"} \\
\text{Assumption: per location } k\in (0,len) \text{ the variances should be the same.} \\
\text{This means that we expect } \eta^n / \eta^t \approx 1

$$
After recess:

$$
\hat{a}_k^{(\#3)} \text{ This is a two-sample estimator that takes GC content into account}\\
\text{This means that we expect } \eta^n / \eta^t \approx 1 \\
\text{Problems with estimators that are a if the denominator is close to 0.} \\
\text{Currently, we will assume that } f_K(GC) \text{ is a constant.}\\
\text{Problems with estimators that are a if the denominator is close to 0.}\\
\\
\hat{a}_k^{(\#3)} = \frac{}{\frac{{Y_k}^n}{{\hat{f}_k}^n{GC}_k}+\mathcal{E}}
$$









Estimator of std. deviation, in regards to the expected mean value.
$$
\frac{\bar{X}_{n_x} - \bar{Z}_{n_z}}
      {\sqrt{
        \frac{\operatorname{Var}(X)}{n_x} +
        \frac{\operatorname{Var}(X)}{n_z} 
      }}
$$
It's problematic to give different weights to X or Z.
It determines what I prefer in regards to the central metric.

In H0: X-Z should be zero.
But even if the pvalue is significant, it's easy to reject 
SNR = Singal to Noise Ratio
Signal = the difference between the expected mean values
Noise = related to the variances that we are investigating


Another way to look at it: 
$$
\Bigg(
\frac{\bar{X}_{n_x} - \bar{Z}_{n_z}}
      {\sqrt{
        \frac{\operatorname{Var}(X)}{n_x} +
        \frac{\operatorname{Var}(X)}{n_z} 
      }}
\Bigg)^2

= \frac{\operatorname{Var}\big[{\operatorname{E}(Y|G=g)}\big]}
       {\operatorname{E}\big[{\operatorname{Var}(Y|G=g)}\big]}{}
$$


```{}
```







