---
title: "Lab Task __"
author: "Group 6"
date: Sys.Date()
output:
  html_document:
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---

Group members:

- Shawn Yakir , 315714980, shawnyakir@gmail.com
- Sarah Levitz, 324673623, sarah.levitz@mail.huji.ac.il
- Carmel Baris, 318455276, carmel.baris@mail.huji.ac.il 


### Libraries used

Please load all libraries that you will use throughout the code in this block.

@TODO
```{r libraries}

#library("")
```


### Paths and Data

Please use this block to write the paths for all the files you are loading. 
If you have a function file, please add it to the submission and call it using a relative path. 

```{r paths}
######## CHANGE THESE FILENAMES
str22_path = ""

```

## Introduction 

Short intro discussing the goal of this week.

## Part 1

Short explanation on method or idea. 


```{r part1, fig.width=5, fig.height=5}

# R code 


# Figure if needed

plot(1:10,1:10, main = "Analysis", xlim= c(1,10), ylim = c(1,10), 
     xlab = "X Variable")


```

Short explanation on what we learn from the graph. 


## Part 2


## Part 3

## Part 4

## Short summary

Something you've learned. 

