---
title: "HW9"
author: "Group 4"
date: "29 July 2024"
output:
  html_document:
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---

Group: 

Group members (Name, ID, email)

- Sigal Falk, 206682031, Sigal.falk@mail.huji.ac.il
- Lihen Laski, 316539774, lihen.laski@mail.huji.ac.il
- Iska Yanir-Amzaleg, 318928587, iska.yanir@mail.huhi.ac.il



### Libraries used

```{r libraries}

library(ggplot2)

if (!require('data.table')){
  install.packages('data.table')
  library('data.table')
}
# install.packages("dplyr", repos = "https://cran.rstudio.com/")
library(dplyr)
library(segmented)

library('splines')
```


### Paths and Data

```{r paths}

load("~/Carmel/RProjects/Lab_Benjamini/data/GC_100.rda")
load("~/Carmel/RProjects/Lab_Benjamini/data/reads_100_A2.rda")
load("~/Carmel/RProjects/Lab_Benjamini/data/reads_100_B2.rda")


short_GC_100 <- GC_100[0:2471993]
# length(short_GC_100)
# length(reads_100_A2)
# length(reads_100_B2)

# Function to merge every 25 cells
merge_25_cells <- function(x) {
  merged <- sapply(seq(1, length(x), by=25), function(i) sum(x[i:min(i+24, length(x))]))
  return(merged)
}

# Apply the function to the datasets (merge every 25 cells)
# Cell of 2500 bases
merged_GC_100 <- merge_25_cells(short_GC_100)
merged_reads_100_A2 <- merge_25_cells(reads_100_A2)
merged_reads_100_B2 <- merge_25_cells(reads_100_B2)
```

### Introduction 

Short intro discussing the goal of this week.


### Preparations

```{r}
remove_outliers_by_std <- function(y, x, num_std) {
  std_y = sd(y)
  mean_y = mean(y)
  
  lower_threshold <- mean_y - num_std*std_y
  upper_threshold <- mean_y + num_std*std_y
  
  non_outlier_indices <- which(y >= lower_threshold & y <= upper_threshold)
  filtered_y <- y[non_outlier_indices]
  filtered_x <- x[non_outlier_indices]
  return(list(filtered_y = filtered_y, filtered_x = filtered_x))
 }


remove_zero_reads <- function(reads, gc) {
  non_zero_indices <- which(reads != 0)
  filtered_reads <- reads[non_zero_indices]
  filtered_gc <- gc[non_zero_indices]
  return(list(filtered_reads = filtered_reads, filtered_gc = filtered_gc))
}


remove_outliers<-function(reads,gc, by_reads= FALSE, by_gc = FALSE, by_zero_reads = FALSE, num_std = 2){
  if (by_reads){
    filtered_data = remove_outliers_by_std(reads,gc,num_std)
    num_outliers_by_reads = length(reads) - length(filtered_data$filtered_y)
    percentage_outliers_by_reads = (num_outliers_by_reads/length(reads))*100
    paste("Percentage of outliers by reads:", percentage_outliers_by_reads)
    reads = filtered_data$filtered_y
    gc = filtered_data$filtered_x
  }
  if (by_gc){
    filtered_data = remove_outliers_by_std(gc,reads,num_std)
    num_outliers_by_reads = length(reads) - length(filtered_data$filtered_x)
    percentage_outliers_by_reads = (num_outliers_by_reads/length(reads))*100
    paste("Percentage of outliers by gc:", percentage_outliers_by_reads)
    reads = filtered_data$filtered_x
    gc = filtered_data$filtered_y
  }
  if (by_zero_reads){
    filtered_data = remove_zero_reads(reads,gc)
    reads = filtered_data$filtered_reads
    gc = filtered_data$filtered_gc
  }
  return(list(filtered_reads = reads, filtered_gc = gc))
}

#T
filtered_data_t = remove_outliers(merged_reads_100_A2,merged_GC_100,by_reads =FALSE, by_gc = FALSE)
filtered_reads_t = filtered_data_t$filtered_reads
filtered_GC_t =  filtered_data_t$filtered_gc

#N
filtered_data_n = remove_outliers(merged_reads_100_B2,merged_GC_100, by_reads =FALSE, by_gc = FALSE)
filtered_reads_n = filtered_data_n$filtered_reads
filtered_GC_n =  filtered_data_n$filtered_gc
```

### Knots selection and Bspline apply
create hat_f g for each sample (cancer and healthy separately)
hat_f_t
hat_f_n
```{r}
bspline_mod <- function(x, y,knots, degree) {
  mod = lm(y~bs(x,knots = knots, degree = degree))
return (mod)
}

knots = c(11,13,16) 
degree=3

model_t = bspline_mod(filtered_GC_t, filtered_reads_t, knots, degree)
model_n = bspline_mod(filtered_GC_n, filtered_reads_n, knots, degree)
```



### Show hat_f_t (for Cancer data)
```{r}
# T
df_t <-data.frame(filtered_GC_t, filtered_reads_t)
resp_t = predict(model_t,df_t)
df_t$resp_t = resp_t
df_t_sorted <- df_t[order(df_t$filtered_GC_t), ]

```

### Show hat_f_n (for Healthy data)
```{r}
# N
df_n <-data.frame(filtered_GC_n, filtered_reads_n)
resp_n = predict(model_n,df_n)
df_n$resp_n = resp_n
df_n_sorted <- df_n[order(df_n$filtered_GC_n), ]

```


### Correction functions
```{r}
# One sample correction with epsilon & No correction (Median correction only)
one_sample_correct = function(samp, preds, eps) {
  a_hat = (samp+eps)/(preds+eps)
  adj_med = 1/median(a_hat)
  return(a_hat*adj_med)
}


# Two sample correction with epsilon (use for both with or without GC correction depends the input vectors)
two_sample_correct_e = function(tumor, normal,eps) {
  a_hat = (tumor+eps)/(normal+eps)
  two_correct_aft_med = a_hat/median(a_hat,na.rm=TRUE)
  return(two_correct_aft_med)
}
```



### One sample correction with epsilon = 1
```{r part1}
eps = 1
corrected_1t = 
  one_sample_correct (df_t$filtered_reads_t,df_t$resp_t,eps)
uncorrected_1t = one_sample_correct(df_t$filtered_reads_t,rep(1,length(df_t$filtered_reads_t)),eps)

corrected_1n =
  one_sample_correct (df_n$filtered_reads_n,df_n$resp_n,eps)
uncorrected_1n = one_sample_correct(df_n$filtered_reads_n,rep(1,length(df_n$filtered_reads_n)),eps)

```


### One sample correction with epsilon = 100
```{r}
eps = 100
corrected_1t = 
  one_sample_correct (df_t$filtered_reads_t,df_t$resp_t,eps)
uncorrected_1t = one_sample_correct(df_t$filtered_reads_t,rep(1,length(df_t$filtered_reads_t)),eps)

corrected_1n =
  one_sample_correct (df_n$filtered_reads_n,df_n$resp_n,eps)
uncorrected_1n = one_sample_correct(df_n$filtered_reads_n,rep(1,length(df_n$filtered_reads_n)),eps)

# find subset
findSubset <- function(start_base, end_base, cell_size){
  cell_start = start_base/cell_size
  cell_end = end_base/cell_size
  chromosome_range = cell_start:cell_end
  return (chromosome_range)
}
```

### Two sample correction with epsilon = 0.1
```{r}

# With eps_1 =100
eps_2 = 0.1
two_sample_with_gc = two_sample_correct_e(corrected_1t,corrected_1n,eps_2)
two_sample_without_gc = two_sample_correct_e(uncorrected_1t,uncorrected_1n,eps_2)

```


### Two sample correction with epsilon = 0.01
```{r}

# With eps_1 =100
eps_2 = 0.01
two_sample_with_gc = two_sample_correct_e(corrected_1t,corrected_1n,eps_2)
two_sample_without_gc = two_sample_correct_e(uncorrected_1t,uncorrected_1n,eps_2)

```

### Defining Ranges A & B
```{r}
# find subset
findSubset <- function(start_base, end_base, cell_size){
  cell_start = start_base/cell_size
  cell_end = end_base/cell_size
  chromosome_range = cell_start:cell_end
  return (chromosome_range)
}


range_a = c(findSubset(25000000,30000000,2500))
#range_0.5_without_gc_outliers = c(1040,1080)

range_b = c(findSubset(28575000,28612500,2500))
 # snr_after <- mean(two_sample_with_gc[range_a]) / sd(two_sample_with_gc[range_b])
 # print(paste("SNR אחרי התיקון:", snr_after))
```

### SNR Calculation

```{r snr_calculate}
x=two_sample_with_gc[range_a]
snr_after1 <- mean(two_sample_with_gc) / sd(two_sample_with_gc[range_a])
print(paste("SNR Only Range A (Post-Fix):", round(snr_after1,1)))
 
m = abs(0.5*mean(two_sample_with_gc[range_b]) - 0.5*mean(two_sample_with_gc[range_a]))
std = sqrt(0.5*var(two_sample_with_gc[range_b]) + 0.5*var(two_sample_with_gc[range_a]))
snr_after2 <- m / std
print(paste("SNR in Range B-A (Post-Fix):", round(snr_after2,1)))

```


### Metrics: MSE, RMSE, and Bais Variance decomposition
```{r}
seperate_vector_to_0.5_1 <- function(vec_correct, range_0.5){
  range_0.5_in_cells = range_0.5+10000
  ind = range_0.5_in_cells[1]:range_0.5_in_cells[2]
  predicted_0.5 = vec_correct[ind]
  predicted_1 = vec_correct[-ind]
  return(list(predicted_0.5=predicted_0.5,predicted_1=predicted_1))
}


mse <- function(predicted, expected_mean) {
  actual = rep(expected_mean, length(predicted))
  paste("For exepected a_hat =",round(expected_mean,3),
        "mse:",round(mean((actual - predicted)^2),3))
}


rmse <- function(predicted, expected_mean) {
  actual = rep(expected_mean, length(predicted))
  paste("For exepected a_hat =", round(expected_mean,3),
        "rmse:", round(sqrt(mean((actual - predicted)^2)),3))
}


bais_var_std <- function(vec_correct, expected_mean){
  bais = mean(vec_correct) - expected_mean
  variance = var(vec_correct)
  std = sd(vec_correct)
  mse_dec = (bais^2) + variance
  paste("For expected a_hat=", expected_mean,"bias:", round(bais,3), 
        "variance:", round(variance,3), "std:", round(std,3), "mse:", round(mse_dec,3))
}
```

### Metrics: Find area of interest (s.t. copy is 0.5) for Calculation
```{r}
range_0.5 = c(1428,1465) # Change if needed
```

### Metrics: Median correction
```{r}
vectors = seperate_vector_to_0.5_1(uncorrected_1t, range_0.5)
vec_0.5 = vectors$predicted_0.5
vec_1 = vectors$predicted_1

mse(vec_0.5,0.5)
rmse(vec_0.5,0.5)
bais_var_std(vec_0.5,0.5)
paste("")
mse(vec_1,1)
rmse(vec_1,1)
bais_var_std(vec_1,1)
```

### Metrics: One Sample correction
```{r}
vectors = seperate_vector_to_0.5_1(corrected_1t, range_0.5)
vec_0.5 = vectors$predicted_0.5
vec_1 = vectors$predicted_1

mse(vec_0.5,0.5)
rmse(vec_0.5,0.5)
bais_var_std(vec_0.5,0.5)
paste("")
mse(vec_1,1)
rmse(vec_1,1)
bais_var_std(vec_1,1)
```

### Metrics: Two Sample correction
```{r}
vectors = seperate_vector_to_0.5_1(two_sample_without_gc, range_0.5)
vec_0.5 = vectors$predicted_0.5
vec_1 = vectors$predicted_1

mse(vec_0.5,0.5)
rmse(vec_0.5,0.5)
bais_var_std(vec_0.5,0.5)
paste("")
mse(vec_1,1)
rmse(vec_1,1)
bais_var_std(vec_1,1)
```

### Metrics: Two Sample + GC correction
```{r}
vectors = seperate_vector_to_0.5_1(two_sample_with_gc, range_0.5)
vec_0.5 = vectors$predicted_0.5
vec_1 = vectors$predicted_1

mse(vec_0.5,0.5)
rmse(vec_0.5,0.5)
bais_var_std(vec_0.5,0.5)
paste("")
mse(vec_1,1)
rmse(vec_1,1)
bais_var_std(vec_1,1)
```



### Plot range A
```{r range_a}
# הגדרת הפרמטרים של הגרף
par(mfcol = c(2,1), mar = c(4, 4, 2, 2))  # מרווחים מותאמים כדי להוסיף כותרות לצירים

# הגדרת התוויות
x_labels <- seq(10000, 12000, by = 500)
num_labels <- length(range_a)
at_positions <- seq(1, num_labels, length.out = length(x_labels))
red_lines_positions <- c(11420, 11470)
red_lines_at <- approx(x = x_labels, y = at_positions, xout = red_lines_positions)$y

# יצירת התרשים הראשי ללא ציר ה-X
plot(two_sample_with_gc[range_a], 
     main = paste("Two Sample Correction with GC 25M:30M"),
     xlab = "Index cell (2500 bases)",  # הוספת כותרת לציר ה-X
     ylab = "Estimated a",  # הוספת כותרת לציר ה-Y
     ylim = c(0, 2),
     cex = 0.5,
     pch = 20,
     xaxt = "n")  # לא להציג את ציר ה-X

# הוספת קו ירוק אופקי
abline(h = 1, col = "green", lw = 2)

# הוספת ציר ה-X עם התוויות העגולות
axis(1, at = at_positions, labels = x_labels)

# הוספת קווים אדומים במיקומים המתאימים
abline(v = red_lines_at, col = "red", lw = 2)

# יצירת התרשים השני ללא ציר ה-X
plot(uncorrected_1t[range_a], 
     main = paste("Median Correction 25M:30M"),
     xlab = "Index cell (2500 bases)",  # הוספת כותרת לציר ה-X
     ylab = "Estimated a",  # הוספת כותרת לציר ה-Y
     ylim = c(0, 2), 
     cex = 0.5, 
     pch = 20,
     xaxt = "n")  # לא להציג את ציר ה-X

# הוספת קו ירוק אופקי
abline(h = 1, col = "green", lw = 2)

# הוספת ציר ה-X עם התוויות העגולות
axis(1, at = at_positions, labels = x_labels)

# הוספת קווים אדומים במיקומים המתאימים
abline(v = red_lines_at, col = "red", lw = 2)
```

### Plot range B
```{r range_b}
plot(two_sample_with_gc[range_b])
plot(uncorrected_1t[range_b])
```

### Hist Trim of Range A
```{r}
trim = function(x,l,u){ pmax(pmin(x, u), l) }
b_range= seq(0,3,0.04)
#use_range = 1:49000
use_range = range_a
par(mfcol = c(2,2))
# hist(trim(uncorrected_1t[use_range],0,3),breaks = b_range,freq = FALSE, main = "Uncorrected")
# hist(trim(corrected_1t[use_range],0,3),breaks = b_range,freq = FALSE,main = "One Sample")
# hist(trim(two_sample_without_gc[use_range],0,3),breaks = b_range,freq = FALSE, main = "Two Sample")
# hist(trim(two_sample_with_gc[use_range],0,3),breaks = b_range,freq = FALSE, main = "Two Sample + GC")

hist(trim(uncorrected_1t[use_range], 0, 3), breaks = b_range, freq = FALSE, main = "Uncorrected", xlab = "index")
hist(trim(corrected_1t[use_range], 0, 3), breaks = b_range, freq = FALSE, main = "One Sample", xlab = "index")
hist(trim(two_sample_without_gc[use_range], 0, 3), breaks = b_range, freq = FALSE, main = "Two Sample", xlab = "index")
hist(trim(two_sample_with_gc[use_range], 0, 3), breaks = b_range, freq = FALSE, main = "Two Sample + GC", xlab = "index")

```



### draft
```{r}
ind = 11070:11100
predicted_0.5 = two_sample_with_gc[11070:11100]
predicted_1 = two_sample_with_gc[-ind]
b_range= seq(0.5,3.5,0.05)
hist(predicted_0.5)
hist(predicted_1)
summary(predicted_1)
summary(predicted_0.5)

sum(predicted_1[predicted_1<(mean(predicted_1)-2*0.09)])/length(predicted_1)
sum(predicted_1[predicted_1>(mean(predicted_1)+2*0.09)])/length(predicted_1)

sum(predicted_0.5[predicted_0.5<(mean(predicted_0.5)-2*sd(predicted_0.5))])/length(predicted_0.5)
sum(predicted_0.5[predicted_0.5>(mean(predicted_0.5)+2*sd(predicted_0.5))])/length(predicted_0.5)

```



