---
title: "Lab Task 8"
author: "Group 6"
date: "`r format(Sys.time(), '%d/%m/%y')`"
output:
  html_document:
    code_folding: show
editor_options: 
  markdown: 
    wrap: sentence
---

Group members:

- Shawn Yakir , 315714980,  shawn.yakir@mail.huji.ac.il
- Sarah Levitz, 324673623, sarah.levitz@mail.huji.ac.il
- Carmel Baris, 318455276, carmel.baris@mail.huji.ac.il 

### Clean the environment
```{r clean0}
rm(list = ls())
```

### Libraries used

```{r libraries}
library('splines')
```


### Paths
```{r paths}
path = "C:/Users/user/Documents/Carmel/Projects/HUJI/Archive_HUJI/data-analysis-genomics/data"
# path = "/Users/alevi/Downloads/"
# path = "\\Users\\shawn\\Desktop\\"
```


### Functions

```{r correcting_funcs}

### Median correction
fix_med = function(samp) {
  return(samp/median(samp,na.rm=TRUE))
}

### One sample correction
fix_gc_e = function(samp, preds, eps=0) {
  return((samp+eps)/(preds+eps))
}


### Two sample correction
fix_2samp_e = function(tumor, healthy, eps=0) {
  return((tumor+eps)/(healthy+eps))
}

```

### Load data
```{r dat}
load(sprintf("%s/reads_100_A2.rda",path))
load(sprintf("%s/reads_100_B2.rda",path))
load(sprintf("%s/GC_100.rda",path))
```

### Binning
```{r binning, warning=FALSE}
binsize = 25

chr2t_mat = matrix(reads_100_A2, nr = binsize)
chr2n_mat = matrix(reads_100_B2, nr = binsize)
GC_100=GC_100[1: (length(reads_100_A2)) ]
gc_mat = matrix(GC_100,nr=binsize)

reads_tumor = colSums(chr2t_mat)
reads_healthy = colSums(chr2n_mat)
GC = colMeans(gc_mat)

```


```{r clean1}
rm("reads_100_A2","reads_100_B2","GC_100","chr2n_mat","chr2t_mat","gc_mat") # remove from memory
```

### Setup for Plots
```{r setup_plots}
# Colors
col_tumor = rgb(.25,.25,.25,0.7)
col_healthy = rgb(0,0,0,0.8)

# Sample1
set.seed(50)
samp1 = sample(length(reads_healthy), 10000)

# Prediction Range
pred_range = seq(0.2,0.75,0.01)

```




### Marginal Distribution of Tumor Reads (Histogram)

```{r hist_tumor}
# Assuming reads_tumor is a vector of read counts per location

# Define breaks for ranges of 100s and group values over 500
breaks <- c(seq(0, 500, by = 100), Inf)
labels <- c(paste(seq(0, 400, by = 100), seq(99, 499, by = 100), sep = "-"), "500+")

# Part 1: Grouping read counts
reads_grouped <- cut(reads_tumor, breaks = breaks, labels = labels, right = FALSE)

# Part 2: Creating a histogram for the grouped data
read_hist <- table(reads_grouped)

# Function to format the values
format_labels <- function(x) {
  if (x >= 1e6) { return(paste0(round(x / 1e6, 2), "M")) }
  if (x >= 1e3) { return(paste0(round(x / 1e3, 1), "K")) } 
  else { return(as.character(x)) }
}

# Define y-axis limits
y_lim <- c(0, max(read_hist) * 1.1)

# Plot the histogram and get the bar midpoints
bar_midpoints <- barplot(read_hist, ylim = y_lim, yaxt = "n", 
                         main = "Tumor Sample",
                         xlab = "Count of Fragments Per Bin", 
                         ylab = "No. of Bins")

# Format y-axis labels
y_ticks <- axTicks(2)
formatted_y_ticks <- sapply(y_ticks, format_labels)
axis(2, at = y_ticks, labels = formatted_y_ticks)

# Format histogram labels
formatted_counts <- sapply(read_hist, format_labels)

# Add formatted labels to the histogram, centered
text(x = bar_midpoints, 
     y = as.numeric(read_hist), 
     labels = formatted_counts, 
     pos = 3, cex = 0.75)
```

### Marginal Distribution of Healthy Reads (Histogram)
```{r hist_healthy}
# Define breaks for ranges of 100s and group values over 500
breaks <- c(seq(0, 500, by = 100), Inf)
labels <- c(paste(seq(0, 400, by = 100), seq(99, 499, by = 100), sep = "-"), "500+")

# Part 1: Grouping read counts
reads_grouped <- cut(reads_healthy, breaks = breaks, labels = labels, right = FALSE)

# Part 2: Creating a histogram for the grouped data
read_hist <- table(reads_grouped)

# Function to format the values
format_labels <- function(x) {
  if (x >= 1e6) { return(paste0(round(x / 1e6, 2), "M")) }
  if (x >= 1e3) { return(paste0(round(x / 1e3, 1), "K")) } 
  else { return(as.character(x)) }
}

# Define y-axis limits
y_lim <- c(0, max(read_hist) * 1.1)

# Plot the histogram and get the bar midpoints
bar_midpoints <- barplot(read_hist, ylim = y_lim, yaxt = "n", 
                         main = "Healthy Sample",
                         xlab = "Count of Fragments Per Bin", 
                         ylab = "No. of Bins")

# Format y-axis labels
y_ticks <- axTicks(2)
formatted_y_ticks <- sapply(y_ticks, format_labels)
axis(2, at = y_ticks, labels = formatted_y_ticks)

# Format histogram labels
formatted_counts <- sapply(read_hist, format_labels)

# Add formatted labels to the histogram, centered
text(x = bar_midpoints, 
     y = as.numeric(read_hist), 
     labels = formatted_counts, 
     pos = 3, cex = 0.75)
```

### Lambda of Poission
```{r poission_lambda}
cat("Median No. of Fragments Per Bin (Tumor): ", median(reads_tumor), "\n")
cat("Mean No. of Fragments Per Bin (Tumor): ", mean(reads_tumor), "\n\n")
cat("Median No. of Fragments Per Bin (Healthy): ", median(reads_healthy), "\n")
cat("Mean No. of Fragments Per Bin (Healthy): ", mean(reads_healthy), "\n")

```


### Spatial Distribution of Healthy vs Tumor Reads
```{r spatial_tumor}

par(mfcol = c(1,1))

x.zoom5=1:length(reads_healthy)
plot(x.zoom5, reads_healthy[x.zoom5],ylim = c(0,400),col=col_healthy, cex =0.5)
lines(loess.smooth(x.zoom5,reads_healthy[x.zoom5], span=0.01),col=4,lw=3)

x.zoom5=1:length(reads_tumor)
plot(x.zoom5, reads_tumor[x.zoom5],ylim = c(0,400),col=col_tumor, cex =0.5)
lines(loess.smooth(x.zoom5,reads_tumor[x.zoom5], span=0.01),col=6,lw=3)
```

```{r spatial_tumor_2.0}

par(mfcol = c(1,1))

x.zoom2=800:2400

plot(x.zoom2, reads_healthy[x.zoom2],ylim = c(0,400),col=col_healthy, cex =0.5)
lines(loess.smooth(x.zoom2, reads_healthy[x.zoom2], span=0.1),col=4,lw=3)

plot(x.zoom2, reads_tumor[x.zoom2],ylim = c(0,400),col=col_tumor, cex =0.5)
lines(loess.smooth(x.zoom2, reads_healthy[x.zoom2], span=0.1),col=4,lw=3)
lines(loess.smooth(x.zoom2, reads_tumor[x.zoom2], span=0.1),col=6,lw=3)
```

### Fragments vs. GC Content (Tumor Sample)

```{r reg_none}
plot(GC[samp1], reads_tumor[samp1], cex = 0.4, pch = 20,  col = col_tumor,
     main = "Tumor Sample vs. GC (%)", ylab = "Fragments Per Bin", xlab = "GC Content",
     ylim = c(0, 400), xlim = c(0.2, 0.75))
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted") # Add gridlines
```


### Linear Regression

```{r reg_lm}
reg_healthy_lm = lm(reads_healthy~GC,subset = reads_healthy>50 & reads_healthy<350)
reg_tumor_lm = lm(reads_tumor~GC,subset = reads_tumor>50 & reads_tumor<400)

resp_healthy_lm = predict(reg_healthy_lm,list(GC=pred_range))
preds_healthy_lm = predict(reg_healthy_lm,list(GC=GC))

response_tumor_lm = predict(reg_tumor_lm,list(GC=pred_range))
preds_tumor_lm = predict(reg_tumor_lm,list(GC=GC))

reg_tumor_lm2 = lm(reads_tumor~ 0 + GC, subset = reads_tumor>50 & reads_tumor<400)
response_tumor_lm2 = predict(reg_tumor_lm2,list(GC=pred_range))

plot(GC[samp1], reads_tumor[samp1], cex = 0.4, pch = 20,  col = col_tumor,
     main = "Tumor Sample", ylab = "Fragments Per Bin", xlab = "GC Content",
     ylim = c(0, 400), xlim = c(0.2, 0.75))
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted") # Add gridlines
lines(pred_range, response_tumor_lm,col = "maroon",lw=3)

summary(reg_tumor_lm)

```

### Broken Sticks Regression

```{r reg_pw}
# Define knots
#myknots <- c(0.415, 0.45, 0.54)
n_knots = 3
myknots = seq(quantile(GC,0.1), quantile(GC,0.9), length.out = n_knots + 2)[-c(1,n_knots + 2)]
pred_range <- seq(0.2, 0.75, 0.01)
# Manual piecewise regression for tumor samples
tumor_segment_1 <- lm(reads_tumor ~ GC, subset = (GC <= myknots[1] & reads_tumor > 50 & reads_tumor < 400))
tumor_segment_2 <- lm(reads_tumor ~ GC, subset = (GC > myknots[1] & GC <= myknots[2] & reads_tumor > 50 & reads_tumor < 400))
tumor_segment_3 <- lm(reads_tumor ~ GC, subset = (GC > myknots[2] & GC <= myknots[3] & reads_tumor > 50 & reads_tumor < 400))
tumor_segment_4 <- lm(reads_tumor ~ GC, subset = (GC > myknots[3] & reads_tumor > 50 & reads_tumor < 400))

# Predict responses for each segment for tumor samples
resp_tumor_segment_1 <- predict(tumor_segment_1, list(GC = pred_range[pred_range <= myknots[1]]))
resp_tumor_segment_2 <- predict(tumor_segment_2, list(GC = pred_range[pred_range > myknots[1] & pred_range <= myknots[2]]))
resp_tumor_segment_3 <- predict(tumor_segment_3, list(GC = pred_range[pred_range > myknots[2] & pred_range <= myknots[3]]))
resp_tumor_segment_4 <- predict(tumor_segment_4, list(GC = pred_range[pred_range > myknots[3]]))

# Plot Tumor Sample with Manual Piecewise Regression
plot(GC[samp1], reads_tumor[samp1], cex = 0.4, pch = 20, col = col_tumor,
     main = "Tumor Sample - Manual Piecewise Regression", ylab = "Fragments Per Bin", xlab = "GC Content",
     ylim = c(0, 400), xlim = c(0.2, 0.75))
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted") # Add gridlines
lines(pred_range[pred_range <= myknots[1]], resp_tumor_segment_1, col = "tan2", lwd = 3)
lines(pred_range[pred_range > myknots[1] & pred_range <= myknots[2]], resp_tumor_segment_2, col = "tan2", lwd = 3)
lines(pred_range[pred_range > myknots[2] & pred_range <= myknots[3]], resp_tumor_segment_3, col = "tan2", lwd = 3)
lines(pred_range[pred_range > myknots[3]], resp_tumor_segment_4, col = "tan2", lwd = 3)
abline(v = myknots, lty = 2, col = "darkblue", lwd = 1)

summary(tumor_segment_1)
summary(tumor_segment_2)
summary(tumor_segment_3)
summary(tumor_segment_4)

```


### B-Spline Regression

```{r reg_bs}
set.seed(50)
samp1 = sample(length(reads_healthy), 10000)

#myknots = c(0.1907, 0.3814, 0.5721)
pred_range = seq(0.2, 0.75, 0.01)

# Fit B-spline models
reg_healthy_bs = lm(reads_healthy ~ bs(GC, degree =3, knots = myknots), subset = reads_healthy > 50 & reads_healthy < 350)
reg_tumor_bs = lm(reads_tumor ~ bs(GC, degree = 3, knots = myknots), subset = (reads_tumor > 50) & (reads_tumor < 400))

# Predict responses for B-spline models
resp_healthy_bs = predict(reg_healthy_bs, list(GC = pred_range))
preds_healthy_bs = predict(reg_healthy_bs, list(GC = GC))

response_tumor_bs = predict(reg_tumor_bs, list(GC = pred_range))
preds_tumor_bs = predict(reg_tumor_bs, list(GC = GC))

# Plot Tumor Sample with B-Spline Regression
plot(GC[samp1], reads_tumor[samp1], cex = 0.4, pch = 20, col = col_tumor,
     main = "Tumor Sample", ylab = "Fragments Per Bin", xlab = "GC Content",
     ylim = c(0, 400), xlim = c(0.2, 0.75))
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted") # Add gridlines
lines(pred_range, response_tumor_bs, col = "darkcyan", lwd = 3)
abline(v = myknots, lty = 2, col = "darkblue", lwd = 1)
summary(reg_tumor_bs)

```

```{r reg_all}
plot(GC[samp1], reads_tumor[samp1], cex = 0.4, pch = 20, col = rgb(0.2,0.2,0.2,0.5),
     main = "Tumor Sample", ylab = "Fragments Per Bin", xlab = "GC Content",
     ylim = c(0, 400), xlim = c(0.2, 0.75))
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted") # Add gridlines

lines(pred_range, response_tumor_lm,col = "maroon",lw=2)
lines(pred_range, response_tumor_bs,col = "darkcyan",lw=3)

lines(pred_range[pred_range <= myknots[1]], resp_tumor_segment_1, col = "tan2", lwd = 3)
lines(pred_range[pred_range > myknots[1] & pred_range <= myknots[2]], resp_tumor_segment_2, col = "tan2", lwd = 3)
lines(pred_range[pred_range > myknots[2] & pred_range <= myknots[3]], resp_tumor_segment_3, col = "tan2", lwd = 3)
lines(pred_range[pred_range > myknots[3]], resp_tumor_segment_4, col = "tan2", lwd = 2)


```

### Applying Estimation Corrections

```{r fixes}
# Median Fix Only (Tumor Sample)
med_eps0 = fix_med(reads_tumor)

# Median & GC Fix (Tumor Sample)
med_gc_1samp_eps0.3 = fix_med(fix_gc_e(reads_tumor, preds_tumor_bs,0.3))

# Median & GC & Two-Sample Fix
gc_fixed_healthy = fix_gc_e(reads_healthy, preds_healthy_bs)
gc_fixed_tumor = fix_gc_e(reads_tumor, preds_tumor_bs)
med_gc_2samp_eps0.3 = fix_med(fix_2samp_e(gc_fixed_tumor, gc_fixed_healthy, 0.3))
```

### Region of Interest

```{r region}
loc2bin = function(v, bin_size){
  return((v*1000000)/bin_size)}

bins = loc2bin(c(25,35,75,80),2500)
stiched_range = c(bins[1]:bins[2], bins[3]:bins[4])
```

### Plotting Fixes in Region of Interest

```{r plot_fixes}
par(mfrow=c(2,1), mar = c(4,4,2,2), oma = c(5,4,3,2))

# Median Fix Only
plot(med_eps0[stiched_range], 
     main = "", ylab = "", xlab = "",
     xaxt = "n",
     ylim = c(0,2), xlim = c(0,6000),
     cex = 0.45, pch=20, col = col_healthy)
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted") # Add gridlines
abline(h=1, col="royalblue", lwd = 3)
abline(v=4000, col="darkblue", lt = 2, lwd = 1.4)
abline(v=c(1420,1470), col="indianred2", lwd = 1.4)
mtext("Median (eps = 0)", side = 3, cex = 1, col = col_tumor, line = 0.5)

# Median & GC & 2Sample
plot(med_gc_2samp_eps0.3[stiched_range], 
     main = "", ylab = "", xlab = "",
     xaxt = "n",
     ylim = c(0,2), xlim = c(0,6000),
     cex = 0.45, pch=20, col = col_healthy)
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted") # Add gridlines
abline(h=1, col="royalblue", lwd = 3)
abline(v=4000, col="darkblue", lt = 2, lwd = 1.4)
abline(v=c(1420,1470), col=6, lwd = 1.4)
mtext("GC & Two-Samp & Median (eps = 0.3)", side = 3, cex = 1, col = col_tumor, line = 0.5)

# Write text on outer margins for the combined plot
mtext("Applying Estimation Fixes", outer = TRUE, side = 3, line = 1, font = 2, cex = 1)
#mtext("Est. Number of Copies (a hat)", outer = TRUE, side = 2, line = 2, cex = 0.8)
mtext("Selected Locations in Chrom1 (25M-35M, 75M-80M)", outer = TRUE, side = 1, line = 2)

# Additional plot to prevent disappearing plots
plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")


```


```{r plot_outliers}
# Save plot to a PNG with increased height
# Adjust height and width in pixels
png("plot_with_increased_height.png", height = 1400, width = 700)

# Basic plot with specified x and y limits
plot(GC, reads_tumor, ylim = c(0, 1400), cex=0.7, pch=20, col=col_tumor)
grid(nx = NULL, ny = NULL, col = "gray", lty = "dotted") # Add gridlines
# Logical condition to identify points within the specified range
non_outliers <- reads_tumor > 50 & reads_tumor < 400
highlight_outside <- !(non_outliers)

# Highlight points outside the specified range
points(GC[highlight_outside], reads_tumor[highlight_outside], col = "indianred2")
abline(h=c(50,400), col="darkblue", lty=2)

# Close the PNG device
dev.off()

```





